Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-31 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: more writing and refs

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det} \, g$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
% \begin{equation} \label{eq:def_gb}
$ \mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
% \end{equation}
%
where $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are respectively the Riemann and Ricci tensors.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
It can however contribute to the field equation when coupled to $\varphi$, as
done through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
%
% Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
% $f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation (see e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GWs
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of black hole binaries (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
%
% These works use the \emph{inspiral} portion of GW event only and a constraint
% on EdGB gravity can be obtained because black holes in this theory carry a
% monopole scalar charge which in turn can source dipolar scalar radiation,
% that affects the GW phase at $-1$PN order, relative to the dominant GR contribution, and
% is proportional to the difference between the charges of binary components.

Second, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
% The calculation of the QNMs in this theory is more complicated than in GR already for
% nonrotating BHs. The reason is due to
The coupling between scalar field and the Gauss-Bonnet invariant, which
manifests into a coupling between scalar perturbation and gravitational
perturbations of polar parity, which, e.g., breaks the familiar equivalence
between the QNM spectra of polar and axial gravitational perturbations (isospectrality) of
Schwarzschild BHs in GR.
%
The QNMs on nonrotating BHs in EdGB gravity were first computed in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations were
extended to leading-order in BH spin in Ref.~\cite{Pierini:2021jxd}.

% \begin{figure}
% \begin{tikzpicture}[baseline,>=stealth]
%     % \draw[help lines, gray] (-4, -2) grid (4, 2);
%     \draw[->] (-2.5, 0) -- (-1.75, +0.5);
%     \draw[->] (-2.5, 0) -- (-1.75, -0.5);
%
%     \node at (-3.25, 0) {$\{h_{\alpha\beta}, \, \varphi\}$};
%
%     \node at (-1, 1.25) {${\cal O}(\chi^{0})$};
%     \node at (-1, +0.5) {$\{ \textcolor{rred}{h^{\rm p}_{\alpha\beta}}, \, \varphi\}$};
%     \node at (-1, -0.5) {$h^{\rm a}_{\alpha\beta}$};
%
%     \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
%     \node at (+1, +0.5) {\dots};
%     \node at (+1, -0.5) {\dots};
% \end{tikzpicture}
% \caption{Quasinormal modes of BHs in scalar-Gauss-Bonnet gravity.}
% \end{figure}

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field, which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ alone is a boundary term~\cite{Jackiw:2003pm},
hence cannot contribute to the field equations in four-dimensions.
%
However, this terms can modify the theory's field equation by coupling it to
$\vartheta$.
%
The coupling strength set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits same the garden-variety Schwarzschild BH solution of
GR~\cite{Jackiw:2003pm}, but admits non-Kerr solutions for spinning BHs~\cite{Yunes:2009hc,Konno:2009kg}.
%
These BHs support a scalar field which asymptotically falls-off as $r^{-2}$, to which we can associate
a scalar dipole charge.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order.
%
Deviations from GR at this PN order are poorly constrained with present GW
observations~\cite{LIGOScientific:2021sio}, making dCS gravity unconstrained by
analyses that rely on the inspiral portion of GWs from BBH
binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref.~\cite{Silva:2020acr}, that found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the x-ray observations of the isolated neutron star (NS)
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star (BNS) GW170817~\cite{TheLIGOScientific:2017qsa}.

@@

% \begin{figure}
% \begin{tikzpicture}[baseline,>=stealth]
%     % \draw[help lines, gray] (-4, -2) grid (4, 2);
%     \draw[->] (-2.5, 0) -- (-1.75, +0.5);
%     \draw[->] (-2.5, 0) -- (-1.75, -0.5);
%
%     \node at (-3.25, 0) {$\{h_{\alpha\beta}, \, \vartheta\}$};
%
%     \node at (-1, 1.25) {${\cal O}(\chi^{0})$};
%     \node at (-1, +0.5) {$h^{\rm p}_{\alpha\beta}$};
%     \node at (-1, -0.5) {$\{ \textcolor{rred}{h^{\rm a}_{\alpha\beta}}, \, \vartheta\}$};
%
%     \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
%     \node at (+1, +0.5) {\dots};
%     \node at (+1, -0.5) {\dots};
% \end{tikzpicture}
% \caption{Quasinormal modes of BHs in dynamical Chern-Simons gravity.}
% \end{figure}

@@

More specifically, we follow Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up
to dimension-eight operators (i.e., $n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    % \nonumber \\
    %         &\quad
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu},
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}},
\label{eq:s_8}
\end{align}
\end{subequations}
%
where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.

@@

% where we defined the Gauss-Bonnet and Pontryagin densities, respectively,
% %
% \begin{subequations}
% \begin{align}
%     R^{2}_{\rm GB} &= R_{\mu\nu\rho\sigma}R^{\mu\nu\rho\sigma} - 4 R_{\mu\nu} R^{\mu\nu} + R^2,
%     \\
%     {}^{*}RR &= \tfrac{1}{2} R_{\nu\mu\rho\sigma} {}^{*}R^{\mu\nu\rho\sigma},
% \end{align}
% \end{subequations}
% %
% and where $R_{\mu\nu\rho\sigma}$ is the Riemann tensor
% and $\tilde{R}_{\mu\nu\rho\sigma}$ its dual, defined as
% %
% \begin{equation}
% \tilde{R}_{\mu\nu\alpha\beta} = \tfrac{1}{2} \epsilon_{\mu\nu\rho\sigma} R^{\rho\sigma}{}_{\alpha\beta},
% \end{equation}
% %
% with $\epsilon_{\mu\nu\rho\sigma}$ being the Levi-Civita tensor.

% This action encapsulates various modified gravity theories, which often are studied on their own,
% namely,
% %
% \begin{itemize}
%     \item scalar Gauss-Bonnet gravity (sGB): corresponds to $\alpha_{\rm GB}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item dynamical Chern-Simons (dCS): this corresponds to $\alpha_{\rm CS}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item cubic effective field theory of GR (cEFTofGR): this corresponds to $\lame$, $\lamo$ nonzero
%           and all other dimensionless coupling constants equal to zero.
%     \item quartic effective field theory of GR (qEFTofGR): this corresponds to $\varepsilon_{i}$
%           ($i = 1,2,3$) nonzero and all other dimensionless coupling constants equal to zero.
% \end{itemize}
%
% In each of these theories, slowly-rotating BH and their quasinormal mode spectra
% to leading order in spin have been calculated.
% %
% We can use this results to establish a mapping between theory-specific calculations and
% the theory-agnostic {\sc Parspec} parametrization.
% %
% We discuss how we do this in the section.
%
% In Eqs.~\eqref{eq:action_mods}, $\alpha_{\rm GB}$, $\alpha_{\rm CS}$, $\lame$,
% $\lamo$ and $\varepsilon_{i}$ are dimensionless constants.

>>>>

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det} \, g$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
%
where $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are respectively the Riemann and Ricci tensors.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
It can however contribute to the field equation when coupled to $\varphi$, as
done through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.

@@

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation (see e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GWs
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of black hole binaries (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.

Second, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
manifests into a coupling between scalar perturbation and gravitational
perturbations of polar parity, which, e.g., breaks the familiar equivalence
between the QNM spectra of polar and axial gravitational perturbations
(isospectrality) of Schwarzschild BHs in GR.
%
The QNMs on nonrotating BHs in EdGB gravity were first computed in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations were
extended to leading-order in BH spin in Ref.~\cite{Pierini:2021jxd}.

@@

This theory is described by the action~\cite{Jackiw:2003pm,Alexander:2009tp},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudoscalar field, which couples to the Pontryagin
density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
As in scalar-Gauss-Bonnet theories, $\int \dV {}^{*}RR$ alone is a boundary
term~\cite{Jackiw:2003pm}, thus does not contribute to the field equations in
four-dimensions.
%
However, this terms can modify the theory's field equation by coupling it to
$\vartheta$.
%
The coupling strength set by the coupling constant $\ell_{\rm CS}$, with
dimensions of length.

The theory admits same the garden-variety Schwarzschild BH solution of
GR~\cite{Jackiw:2003pm}, but admits non-Kerr solutions for spinning BHs~\cite{Yunes:2009hc,Konno:2009kg}.
%
These BHs support a scalar field which asymptotically falls-off as $r^{-2}$, to which we can associate
a scalar dipole charge.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order~\cite{Yagi:2011xp,Alexander:2017jmt}.
%
Deviations from GR at this PN order are poorly constrained with present GW
observations~\cite{LIGOScientific:2021sio}, making dCS gravity unconstrained by
analyses that rely on the inspiral portion of GWs from BBH
binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref.~\cite{Silva:2020acr}, that found
%
$\ell_{\rm CS} \leqslant 8.5$~km,
%
by folding data from the x-ray observations of the isolated neutron star (NS)
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star (BNS) GW170817~\cite{TheLIGOScientific:2017qsa}.

@@

@@

More specifically, we follow Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up
to dimension-eight operators (i.e., $n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu},
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}},
\label{eq:s_8}
\end{align}
\end{subequations}
%
where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.

@@
